---
title: "Creating a custom steps function"
author: "Casey Visintin"
date: "`r Sys.Date()`"
output:
  pdf_document
geometry: margin=2cm
vignette: >
  %\VignetteIndexEntry{Custom Function}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  dpi = 300,
  fig.width = 7,
  out.width = "100%",
  cache = TRUE
)
```

Perhaps one of the most useful features of *steps* is its flexibility. Users have the options of selecting pre-defined functions or writing their own functions to use in simulations.  This vignette will describe how to implement custom functions in the *steps* software.

First, we load required packages:

```{r, message = FALSE}
library(steps)
library(raster)
library(viridisLite)
library(future)
```

To better facilitate the use of custom functions, a basic understanding of the *steps* internal structure is required. We have designed a simple and straight-forward application programming interface (API). In its most basic operation, *steps* passes landscape objects and iterator values (timesteps) to functions. These functions then return landscape objects that have been modified based on operations specified in the functions:

\begin{center}
\includegraphics[width=370px]{steps_diagram.png}
\end{center}

Landscape objects contain information on populations, carrying capacities, habitat features, and other spatially-explicit information used by dynamic functions in a simulation. Dynamics functions act on these landscape objects during a simulation.

Let's examine a pre-built *steps* dynamics function "mortality" which changes populations based on spatially-explicit values provided in a raster layer.

```{r, message = FALSE}

mortality <- function (mortality_layer, stages = NULL) {
  
  pop_dynamics <- function (landscape, timestep) {

      population_raster <- landscape$population
      nstages <- raster::nlayers(population_raster)
      
      # get population as a matrix
      idx <- which(!is.na(raster::getValues(population_raster[[1]])))
      population_matrix <- raster::extract(population_raster, idx)
      
      mortality_prop <- raster::extract(landscape[[mortality_layer]][[timestep]], idx)
      
      if (is.null(stages)) stages <- seq_len(nstages)
      
      for (stage in stages) {
        
        population_matrix[ , stage] <- ceiling(population_matrix[ , stage] * mortality_prop)
        
      }
      
      # put back in the raster
      population_raster[idx] <- population_matrix
      
      landscape$population <- population_raster

    
    landscape
    
  }
  
  as.population_modification(pop_dynamics)
  
}
```    

In the first line, we have defined a function with parameters that specify the name of the layer in the landscape object that will be used to modify populations (mortality_layer) and which life stages will be modified (stages). We have set "stages" to NULL as a default so, if not specified, all stages will be modified. These are global parameters for the function and are optional.

We define an internal function (returned by the main function) on the second line and this is where the dynamic operations are specified. Note the two parameters are "landscape" and "timestep" and these are both required.

The next few lines operate on the landscape object by extracting the populations of the affected stages, extracting the values of the "mortality layer", multiplying them to get new populations, and then returning the modified populations to the landscape object. Note the last line is always "landscape" so the landscape object is returned when the function concludes its operations.

The final line assigns a class to the function. This is optional but useful to maintain an organised structure in the software. We recommend using the appropriate function based on the category of the function a user is defining. In this case, this function is a "population_modification" class since its operations change the population in the landscape object.

So let's go about creating a custom function for use in *steps*. First, we load some initial kangaroo populations:

```{r, message = FALSE, fig.align = "center"}
egk_pop

par(mar=c(0,0,0,0), oma=c(0,0,0,0))
spplot(egk_pop, col.regions = viridis(100))
```

Let's examine the total population of juveniles:

```{r, message = FALSE}
cellStats(egk_pop[[1]], sum)
```

Now, let's create a function that reduces cell populations randomly across the landscape by a given percentage for the life stages specified. We start with the first two lines coded similar to the previous example and change the function name and one of the parameters:

```{r, message = FALSE, eval = FALSE}
percent_mortality <- function (percentage, stages = NULL) {
  
  pop_dynamics <- function (landscape, timestep) { ...
```

Next we extract the populations similar to the previous example. We use an index to avoid extracting missing values:

```{r, message = FALSE, eval = FALSE}
percent_mortality <- function (percentage, stages = NULL) {
  
  pop_dynamics <- function (landscape, timestep) {
      population_raster <- landscape$population
      nstages <- raster::nlayers(population_raster)
      
      # get population as a matrix
      idx <- which(!is.na(raster::getValues(population_raster[[1]])))
      population_matrix <- raster::extract(population_raster, idx) ...
```

We then randomly sample cells (by non-zero indices) to be modified through an iterative process and then stop when we reach our specified target percentage reduction:

```{r, message = FALSE, eval = FALSE}
percent_mortality <- function (percentage, stages = NULL) {
  
  pop_dynamics <- function (landscape, timestep) {
      population_raster <- landscape$population
      nstages <- raster::nlayers(population_raster)
      
      # get population as a matrix
      idx <- which(!is.na(raster::getValues(population_raster[[1]])))
      population_matrix <- raster::extract(population_raster, idx)
      
      if (is.null(stages)) stages <- seq_len(nstages)
      for (stage in stages) {
        initial_pop <- sum(population_matrix[ , stage])
        changing_pop <- sum(population_matrix[ , stage])
        while (changing_pop > initial_pop * percentage) {
          non_zero <- which(population_matrix[idx , stage] > 0)
          i <- sample(non_zero, 1)
          population_matrix[i , stage] <- population_matrix[i , stage] - 1
          changing_pop <- sum(population_matrix[ , stage])
        }
      } ...
```

Finally, similar to the original example, we return the modified populations to the landscape object and then return it from the function:

```{r, message = FALSE}
percent_mortality <- function (percentage, stages = NULL) {
  
  pop_dynamics <- function (landscape, timestep) {
    population_raster <- landscape$population
    nstages <- raster::nlayers(population_raster)
    
    # get population as a matrix
    idx <- which(!is.na(raster::getValues(population_raster[[1]])))
    population_matrix <- raster::extract(population_raster, idx)
    
    if (is.null(stages)) stages <- seq_len(nstages)
    for (stage in stages) {
      initial_pop <- sum(population_matrix[ , stage])
      changing_pop <- sum(population_matrix[ , stage])
      while (changing_pop > initial_pop * (1 - percentage)) {
        non_zero <- which(population_matrix[ , stage] > 0)
        i <- sample(non_zero, 1)
        population_matrix[i , stage] <- population_matrix[i , stage] - 1
        changing_pop <- sum(population_matrix[ , stage])
      }
    }
    
    population_raster[idx] <- population_matrix
    landscape$population <- population_raster
    landscape
  }
} 
```

Now that we have a custom function defined in the R environment we can specify it in the model setup. Note, we specify a 10% (percentage = 0.1) reduction in the total landscape population of juveniles (stages = 1) and run the simulation for a single replicate of three iterations (timesteps = 3):

```{r, message = FALSE}
ls <- landscape(population = egk_pop,
                suitability = NULL,
                carrying_capacity = NULL)

pd <- population_dynamics(change = NULL,
                          dispersal = NULL,
                          modification = percent_mortality(percentage = 0.1,
                                                           stages = 1),
                          density_dependence = NULL)

results <- simulation(landscape = ls,
                      population_dynamics = pd,
                      habitat_dynamics = NULL,
                      demo_stochasticity = "none",
                      timesteps = 3,
                      replicates = 1,
                      verbose = FALSE)
```

We can view the initial total juvenile landscape population and the modified population in each of the timesteps:

```{r, message = FALSE}
cbind(c("t0", "t1", "t2", "t3"),
      c(cellStats(egk_pop[[1]], sum),
        cellStats(results[[1]][[1]][["population"]][[1]], sum),
        cellStats(results[[1]][[2]][["population"]][[1]], sum),
        cellStats(results[[1]][[3]][["population"]][[1]], sum)))

```
And here is a plot of the temporally-changing juvenile cell populations:

```{r, message = FALSE, fig.align = "center"}
pop_stack <- stack(egk_pop[[1]],
                   results[[1]][[1]][["population"]][[1]],
                   results[[1]][[2]][["population"]][[1]],
                   results[[1]][[3]][["population"]][[1]])

names(pop_stack) <- c("t0", "t1", "t2", "t3")

par(mar=c(0,0,0,0), oma=c(0,0,0,0))
spplot(pop_stack, col.regions = viridis(100))
```

Because of the internal organisation and structure of *steps*, we are also able to use any information contained within the landscape object in our custom functions. For example, say we would like to reduce the populations in only highly suitable habitat (above a threshold value) - this can be parameterised in the function. 

We add an additional parameter to specify a threshold value indicating which cells in the landscape that we will randomly reduce populations in. To do this, we must also extract information from the suitability layer in the landscape object:

```{r, message = FALSE}
percent_mortality_hab <- function (percentage, threshold, stages = NULL) {
  
  pop_dynamics <- function (landscape, timestep) {
    population_raster <- landscape$population
    nstages <- raster::nlayers(population_raster)
    
    # get population as a matrix
    idx <- which(!is.na(raster::getValues(population_raster[[1]])))
    population_matrix <- raster::extract(population_raster, idx)
    
    # get suitability cell indexes
    suitable <- raster::extract(landscape$suitability, idx)
    
    if (is.null(stages)) stages <- seq_len(nstages)
    for (stage in stages) {
      initial_pop <- sum(population_matrix[ , stage])
      changing_pop <- sum(population_matrix[ , stage])
      
      highly_suitable <- which(suitable >= threshold) # check for suitable cells
      
      while (changing_pop > initial_pop * (1 - percentage)) {
        non_zero <- which(population_matrix[ , stage] > 0)
        i <- sample(intersect(highly_suitable, non_zero), 1) # change the sampling pool
        population_matrix[i , stage] <- population_matrix[i , stage] - 1
        changing_pop <- sum(population_matrix[ , stage])
      }
    }
    
    population_raster[idx] <- population_matrix
    landscape$population <- population_raster
    landscape
  }
} 
```

Now we run our simulation again using the new custom function and a suitability threshold of 0.7. Note, we must also now include a suitability raster layer in the landscape object or else our function will not work:

```{r, message = FALSE}
ls <- landscape(population = egk_pop,
                suitability = egk_hab,
                carrying_capacity = NULL)

pd <- population_dynamics(change = NULL,
                          dispersal = NULL,
                          modification = percent_mortality_hab(percentage = .10,
                                                               threshold = 0.7,
                                                               stages = 1),
                          density_dependence = NULL)

results <- simulation(landscape = ls,
                      population_dynamics = pd,
                      habitat_dynamics = NULL,
                      demo_stochasticity = "none",
                      timesteps = 3,
                      replicates = 1,
                      verbose = FALSE)
```

Again, we can view the initial total juvenile landscape population and the modified population in each of the timesteps:

```{r, message = FALSE}
cbind(c("t0", "t1", "t2", "t3"),
      c(cellStats(egk_pop[[1]], sum),
        cellStats(results[[1]][[1]][["population"]][[1]], sum),
        cellStats(results[[1]][[2]][["population"]][[1]], sum),
        cellStats(results[[1]][[3]][["population"]][[1]], sum)))

```
And here is an updated plot of the juvenile cell populations:

```{r, message = FALSE, fig.align = "center"}
pop_stack <- stack(egk_pop[[1]],
                   results[[1]][[1]][["population"]][[1]],
                   results[[1]][[2]][["population"]][[1]],
                   results[[1]][[3]][["population"]][[1]])

names(pop_stack) <- c("t0", "t1", "t2", "t3")

par(mar=c(0,0,0,0), oma=c(0,0,0,0))
spplot(pop_stack, col.regions = viridis(100))
```

Notice that whilst the total juvenile population numbers have decreased in a similar fashion, the spatial locations of the reductions have been confined to areas where habitat is highly suitable - as specified by our threshold value.

But what if we want to influence the locations of the random samples by our own custom raster layer? This is accomplished by creating/loading a new raster that matches the properties (extent, resolution, and projection) of the population raster stack and adding it to the landscape object:

```{r, message = FALSE, fig.align = "center"}
sampling_mask <- egk_hab # copy the habitat layer to assume its properties
sampling_mask[!is.na(egk_hab)] <- 0 # set all non-NA values to zero

idx_corner <- sort(unlist(lapply(0:15, function (x) seq(21, 541, by = 36) + x))) # get index values of the top-right corner (15 x 15 cells)
sampling_mask[intersect(idx_corner, which(!is.na(egk_hab[])))] <- 1 # set non-NA raster values in top-right corner to one

plot(sampling_mask)

ls <- landscape(population = egk_pop,
                suitability = NULL,
                carrying_capacity = NULL,
                "sampling_mask" = sampling_mask) # name the object to reference it in our function
```

We now need to slightly alter our custom function to accept and use the new raster. We start with the same syntax, adding a parameter to specify the name of the raster in the landscape object to use in our function and then extracting using the specified name:

```{r, message = FALSE}
percent_mortality_ras <- function (percentage, custom_raster, threshold, stages = NULL) {
  
  pop_dynamics <- function (landscape, timestep) {
    population_raster <- landscape$population
    nstages <- raster::nlayers(population_raster)
    
    # get population as a matrix
    idx <- which(!is.na(raster::getValues(population_raster[[1]])))
    population_matrix <- raster::extract(population_raster, idx)
    
    # Note, here we now use the name of the raster ('custom_raster' parameter)
    # to extract the object from the landscape object
    suitable <- raster::extract(landscape[[custom_raster]], idx)
    
    if (is.null(stages)) stages <- seq_len(nstages)
    for (stage in stages) {
      initial_pop <- sum(population_matrix[ , stage])
      changing_pop <- sum(population_matrix[ , stage])
      
      highly_suitable <- which(suitable >= threshold) # check for suitable cells
      
      while (changing_pop > initial_pop * (1 - percentage)) {
        non_zero <- which(population_matrix[ , stage] > 0)
        i <- sample(intersect(highly_suitable, non_zero), 1) # change the sampling pool
        population_matrix[i , stage] <- population_matrix[i , stage] - 1
        changing_pop <- sum(population_matrix[ , stage])
      }
    }
    
    population_raster[idx] <- population_matrix
    landscape$population <- population_raster
    landscape
  }
} 
```

In the 'population_dynamics' function, we change the modification parameter to call the new custom function (percent_mortality_ras) and add the new parameter specifying the name of the raster in the landscape object. Since we used zeros and ones in our sampling mask it does not matter which value we select for the threshold - as long as it is between zero and one. As previously, we run the simulation for a single replicate of three iterations:

```{r, message = FALSE}
pd <- population_dynamics(change = NULL,
                          dispersal = NULL,
                          modification = percent_mortality_ras(percentage = .10,
                                                               custom_raster = "sampling_mask",
                                                               threshold = 0.7,
                                                               stages = 1),
                          density_dependence = NULL)

results <- simulation(landscape = ls,
                      population_dynamics = pd,
                      habitat_dynamics = NULL,
                      demo_stochasticity = "none",
                      timesteps = 3,
                      replicates = 1,
                      verbose = FALSE)
```

The changes in total juvenile landscape population are as expected:

```{r, message = FALSE}
cbind(c("t0", "t1", "t2", "t3"),
      c(cellStats(egk_pop[[1]], sum),
        cellStats(results[[1]][[1]][["population"]][[1]], sum),
        cellStats(results[[1]][[2]][["population"]][[1]], sum),
        cellStats(results[[1]][[3]][["population"]][[1]], sum)))

```

But now the changes in juvenile cell populations are confined to our 'sampling_mask' layer:

```{r, message = FALSE, fig.align = "center"}
pop_stack <- stack(egk_pop[[1]],
                   results[[1]][[1]][["population"]][[1]],
                   results[[1]][[2]][["population"]][[1]],
                   results[[1]][[3]][["population"]][[1]])

names(pop_stack) <- c("t0", "t1", "t2", "t3")

par(mar=c(0,0,0,0), oma=c(0,0,0,0))
spplot(pop_stack, col.regions = viridis(100))
```

So far we have been modifying the population rasters in the landscape object, but we can also modify transition matrices as well. This may be useful in situations where a user would like to deterministically define, or affect, survival and fecundity.

In the next example, we demonstrate the use of rasters to modify the survival of juvenile kangaroos. We begin with a simple simulation of population growth. For this simulation we only need initial populations and a growth function based on a transition matrix - environmental stochasticity is set to zero by default in the function. We turn of demographic stochasticity and run the simulation for a single replicate of five iterations:

```{r, message = FALSE}
ls <- landscape(population = egk_pop,
                suitability = NULL,
                carrying_capacity = NULL)

pd <- population_dynamics(change = growth(transition_matrix = egk_mat),
                          dispersal = NULL,
                          modification = NULL,
                          density_dependence = NULL)

results <- simulation(landscape = ls,
                      population_dynamics = pd,
                      habitat_dynamics = NULL,
                      demo_stochasticity = "none",
                      timesteps = 5,
                      replicates = 1,
                      verbose = FALSE)
```

Let's look at how the total juvenile population changes through time:

```{r, message = FALSE, fig.align = "center"}
plot(results, stages = 1)
```

And here is the spatial distribution of cell populations for juvenile kangaroos through time:

```{r, message = FALSE, fig.align = "center"}
plot(results,
     stages = 1,
     type = "raster",
     timesteps = 1:5,
     panels = c(5, 1))
```

If we look at the transition matrix used in the simulation, we can see why there is a steady increase in population (Rmax ~ 1.17):

```{r, message = FALSE}
print(egk_mat)

paste("Rmax =", abs(eigen(egk_mat)$values[1]))

```

Now let's suppose that we want to provide rasters that influence the survival of juvenile kangaroos. The methods are quite similar to our earlier examples but we will need to use a bit more indexing. First, we will construct a raster that will contain our values of influence - in this case numbers between zero and one that will be multiplied with juvenile survival in each timestep:

```{r, message = FALSE, fig.align = "center"}
survival_influence <- egk_hab # copy the habitat layer to assume its properties
survival_influence[!is.na(egk_hab)] <- 1 # set all non-NA values to one

idx_corner <- sort(unlist(lapply(0:15, function (x) seq(21, 541, by = 36) + x))) # get index values of the top-right corner (15 x 15 cells)
idx_corner <- intersect(idx_corner, which(!is.na(egk_hab[]))) # remove NA cells in index

survival_influence[intersect(idx_corner, which(!is.na(egk_hab[])))] <- runif(length(idx_corner)) # assign random numbers between 0 and 1 to indexed cells 

plot(survival_influence)

```


These are simple examples, but users can specify a seemingly endless configuration of dynamics to be used throughout a simulation by following the structure and methods introduced in this vignette. Nnything can be stored in a landscape object and then be referred to within a custom function. Thus, the procedure for creating custom functions also applies to habitat dynamics - which are called in the simulation function.

We encourage users to examine the pre-built functions in *steps* to get an idea of their internal operations. Further, these pre-built functions can serve as templates to get users started in customising and creating their own functions.
